---
title: "BI 328 Conservation Biology"
author: "Labmanual Fall 2022"
date: "`r Sys.Date()`"
knit: "bookdown::preview_chapter"
output:
  msmbstyle::msmb_html_book:
    highlight: tango
    toc: TRUE
    toc_depth: 1
    split_by: chapter
    margin_references: FALSE
    css: msmb.css
bibliography: labmanual.bib
link-citations: yes
editor_options: 
  chunk_output_type: console
---

```{r include=FALSE}


library(msmbstyle)

# invalidate cache when the tufte version changes
knitr::opts_chunk$set(tidy = FALSE, 
                      message = FALSE,
                      error = FALSE,
                      warning = FALSE,
                      cache.extra = packageVersion('msmbstyle'))


options(htmltools.dir.version = FALSE)

```

# Trees vs Albedo






```{r}

# load libraries
library(tidyverse)
library(knitr)
library(janitor)

# load custom ggplot themes
source("scr/ggplot.R")

```

## Central Question

[Make changes requested in your google doc and add here]


## Hypothesis

[Make changes requested in your google doc and add here]


## Objectives

[Make changes requested in your google doc and add here]


## Methods

### Study Design

[Make changes requested in your google doc and add here]

### Data Collection and Processing

[Make changes requested in your google doc and add here]

### Data Analysis




## Author Contributions

Typical categories include: Conceive study, design study, project/task management, data collection, data management, data analysis, visualizations, write/design manuscript/poster draft, revise & edit draft.

Write about 5-7 sentences about how you contributed to this project and what you learned about the process from the wibbly-wobbly method of just diving right in and designing your own study this semester.


## If I (we) were to design this study all over again ...

Find at least three points for improvement/lessons learned about the study design itself (separate from what you learned about the process). Discuss this with your team members but you should write your own revisions in about 7-10 sentences.


## Further work: Next steps and new questions based on our pilot study ...

Discuss with your team mates to identify new questions you have identified, how you might expand and refine the study (beyond just re-designing the pilot study), assume that you are not limited by the constraints of lab time within a course.








